### 2024-09-02 chap_atom_mol/chap_atom_mol.tex
Ready to print

\subsection{Power sector}
\label{subsec:atom_mol:results_deter_power_sector}
\gls{SMR} is deployed as soon as available, \ie 2040, to its maximum capacity, 6\,GW, substituting other flexible power generation units (see \Cref{fig:results_deter_tech_cap_elec}). There is no ammonia-\gls{CCGT} at the end of the transition and an anticipatory reduction of methane \gls{CCGT} (\ie 2.1\,GW in 2040 versus 3.7\,GW for the REF case). To a lesser extent, the last 2\% deployment of solar-\gls{PV} is slightly delayed as the capacity in 2025 is 1.3\,GW smaller than in the REF case. Overall, given the restriction on yearly availability and the slightly higher electrification (\Cref{fig:results_deter_layer_elec}), the total power capacity installed by 2050 is 3.5\% higher for the SMR case. In their work, \citet{PATHS2050} also showed that \gls{SMR} first substitutes ``e-fuels/hydrogen'' turbines before reducing the need for \gls{PV}. However, in their ``Central'' scenario where no \gls{SMR} is installed by 2050, they rely on 96.6\,GW of \gls{PV}, 63\% more than the 59.2\,GW potential considered in our work and about 15 times more than the current installed capacity, \ie 6.5\,GW. \\

\begin{figure}[htbp!]
\centering
\includegraphics[width=\textwidth]{Elec_Tech_Cap_REF.pdf}
\includegraphics[width=\textwidth]{Elec_Tech_Cap_SMR.pdf}
\caption{Installed capacities in the power sector in the REF (top) and SMR (down) scenarios. As soon as available (\ie 2040), \acrfull{SMR} is deployed to its maximum potential (\ie 6\,GW) to substitute more expensive flexible generation units (\ie methane and ammonia \gls{CCGT}).}
\label{fig:results_deter_tech_cap_elec}
\end{figure}

\newpage
When assessing the electricity production-versus-consumption-balance (\Cref{fig:results_deter_layer_elec}), \gls{SMR}, as a cheaper, flexible and low-emitting power generation system, produces to its full capacity. Given the 15\% maintenance off-time assumed in this work, this represents 44.6\,TWh by 2050. In comparison, in 2020, conventional nuclear power plants produced 34.4\,TWh in Belgium. This resurgence of nuclear electricity occurs at the expense of other, although more efficient, technologies: \gls{CCGT} and industrial \gls{CHP}. Besides the unchanged end-use-demand compared with the REF case, we observe a slight increase in the electrification of the rest of the system: +9.4\% which corresponds to +5.8\,TWh, mostly consumed by electric heaters in industry (+48\%) to produce industrial high-temperature heat.

\begin{figure}[htbp!]
\centering
\includegraphics[width=\textwidth]{Elec_Layer_REF.pdf}
\includegraphics[width=\textwidth]{Elec_Layer_SMR.pdf}
\caption{Balance between the production and the consumption of electricity in the REF (top) and SMR (down) scenarios. The production of electricity from \gls{SMR} substitutes more efficient technologies (\ie \gls{CCGT} and \gls{CHP}) and boosts the electrification of the rest of the system, mostly the industrial high-temperature heat sector.}
\label{fig:results_deter_layer_elec}
\end{figure}

@@

Then, considering the primary energy mix shown in \Cref{fig:results_deter_energy_mix}, three phases in the transition can be identified. Before 2040, thanks to the perfect foresight approach, the model finds it more economical in 2025 to keep on using 33.2\,TWh of \gls{LFO} to produce \gls{HVC} through naphtha/LPG-cracking. In 2040, uranium-driven \gls{SMR} substitutes the electricity originally produced from industrial \gls{CHP} and \gls{CCGT} running respectively on fossil gas and renewable ammonia. Finally, from 2045 onward, the significant drop in the consumption of electrofuels comes from the same industrial \gls{CHP}. This is the illustration of the atom-molecules dilemma where the consumption of local renewables is, on its side, not much affected. In other words, \gls{SMR} competes with importing electrofuels while both support the integration of local solar and wind energies. Then, like the power sector, the SMR case ends up in a less efficient whole-energy system by 2050 at it consumes 47\,TWh (+12.7\%) primary energy more to supply an unchanged \gls{EUD}. Interestingly, in both cases, given the assumptions made on the \gls{GWP} of the resources (\ie $\mathit{gwp}_{\mathrm{op,electrofuels}}=0$\,kt$_{\ce{CO2},\text{eq}}$/GWh and $\mathit{gwp}_{\mathrm{op,uranium}}=0.004$\,kt$_{\ce{CO2},\text{eq}}$/GWh), the constraint on the \ce{CO2} budget leads to ``carbon neutrality'' by 2050\footnote{The model being constrained to keep on using all the waste that would keep on being locally produced, the system in 2050 reaches a $\sim 3.5\,Mt_{\ce{CO2},\text{eq}}$/year.}.

@@

As aforementioned, the industrial \gls{EUD} impacts the most the import of e-methane (Figure \ref{fig:results_uq_samples_gas}). This parameter directly dictates the demand of industrial high-temperature heat for which industrial gas \gls{CHP}, and industrial gas boilers to a lower extent, represent, on average over all the samples, respectively 25.6\% and 6.1\% of the total production. Then, considering the smaller-impact parameters, we notice that \gls{SMR} plays a non-negligible role. Indeed, if deployed, \gls{SMR} produces abundant low-emitting electricity for industrial electric heaters that substitute, even completely in some cases, gas alternatives. This confirms the observation made in Section \ref{subsec:atom_mol:results_deter_others}. Highly available local biomass also leads to smaller imports of e-methane to supply bio-hydrolysis and produce methane-equivalent gas. Finally, and surprisingly, the costs of purchasing electrofuels and fossil fuels have a positive and negative correlation with the amount of e-methane, respectively. In other words, by 2050, more expensive electrofuels result in more e-methane imports. On the contrary, cheaper electrofuels result in higher imports of fossil methane. Given the techno-economic optimum sought by EnergyScope, if electrofuels are more expensive, the system will, overall, import less of them, especially e-ammonia, mainly used by \gls{CCGT}. Subject to the \ce{CO2} budget for the transition, the system goes towards more efficient technologies, like industrial methane-\gls{CHP} to substitute e-ammonia-\gls{CCGT} in the production of electricity. First running on fossil gas, these \gls{CHP} consume more e-methane by 2050. On the contrary, if electrofuels are cheaper, there is more import of them, especially of e-ammonia. This leaves room for more emitting and cheaper resources to be used while respecting the \ce{CO2} budget, \ie coal in industrial boilers that produce, in these cases, on average 24\% of the high-temperature (HT) heat in 2050. In these cases, the use of coal in 2050 highlights that, with a sharper decrease of the emissions at early stages in the transition, the model finds a solution including highly-emitting resources (\eg coal) while respecting the \ce{CO2} budget. Consequently, there are smaller investments in methane \gls{CHP}, and consequently import of e-methane as more abundant renewable electricity is produced via e-ammonia-\gls{CCGT} and more HT-heat is supplied by industrial coal boilers. Even though we might expect that no more coal will be consumed in Belgium by 2050, the model still has the opportunity to use it if the \ce{CO2} budget allows it. Regarding the cost of purchasing fossil fuels, the parameter has mainly an impact on the import of fossil \gls{NG} as the most versatile energy carrier in the whole-energy system. If \gls{NG} is more expensive, the system will import less of it. Subsequently, the investments in methane-\gls{CHP} and boilers are more limited. This ends up in a smaller need for e-methane by 2050.

@@

@@

\section{Conclusions}
\label{sec:atom_mol:conclusions}
Given its lower \gls{LCOE} (40\,€/MWh$_{\text{elec}}$) and the important share of the investment recovered via its salvage value (between 79\% and 96\%), \gls{SMR} is installed as soon as available. It directly substitutes other flexible power generation units (\ie CCGT) and provides, by 2050, 44.6\,TWh, 25\% of the total electricity production. Consequently, this reduces the need to produce electricity via \gls{CHP} and allows increasing by 48\% the high-temperature heat produced by electric heaters. Besides these two sectors, the others (\ie low-temperature heat, mobility and non-energy demand) are marginally or not impacted by the integration of \gls{SMR}.

>>>>

\subsection{Power sector}
\label{subsec:atom_mol:results_deter_power_sector}
\gls{SMR} is deployed as soon as available, \ie 2040, to its maximum capacity, 6\,GW, substituting other flexible power generation units (see \Cref{fig:results_deter_tech_cap_elec}). There is no ammonia-\gls{CCGT} at the end of the transition and an anticipatory reduction of methane \gls{CCGT} (\ie 2.1\,GW in 2040 versus 3.7\,GW for the REF case). To a lesser extent, the last 2\% deployment of solar-\gls{PV} is slightly delayed as the capacity in 2025 is 1.3\,GW smaller than in the REF case. Overall, given the restriction on yearly availability and the slightly higher electrification (\Cref{fig:results_deter_layer_elec}), the total power capacity installed by 2050 is 3.5\% higher for the SMR case. In their work, \citet{PATHS2050} also showed that \gls{SMR} first substitutes ``e-fuels/hydrogen'' turbines before reducing the need for \gls{PV}. However, in their ``Central'' scenario where no \gls{SMR} is installed by 2050, they rely on 96.6\,GW of \gls{PV}, 63\% more than the 59.2\,GW potential considered in our work and about 15 times more than the current installed capacity, \ie 6.5\,GW.

\begin{figure}[t!]
\centering
\includegraphics[width=\textwidth]{Elec_Tech_Cap_REF.pdf}
\includegraphics[width=\textwidth]{Elec_Tech_Cap_SMR.pdf}
\caption{Installed capacities in the power sector in the REF (top) and SMR (down) scenarios. As soon as available (\ie 2040), \acrfull{SMR} is deployed to its maximum potential (\ie 6\,GW) to substitute more expensive flexible generation units (\ie methane and ammonia \gls{CCGT}).}
\label{fig:results_deter_tech_cap_elec}
\end{figure}

When assessing the electricity production-versus-consumption-balance (\Cref{fig:results_deter_layer_elec}), \gls{SMR}, as a cheaper, flexible and low-emitting power generation system, produces to its full capacity. Given the 15\% maintenance off-time assumed in this work, this represents 44.6\,TWh by 2050. In comparison, in 2020, conventional nuclear power plants produced 34.4\,TWh in Belgium. This resurgence of nuclear electricity occurs at the expense of other, although more efficient, technologies: \gls{CCGT} and industrial \gls{CHP}. Besides the unchanged end-use-demand compared with the REF case, we observe a slight increase in the electrification of the rest of the system: +9.4\% which corresponds to +5.8\,TWh, mostly consumed by electric heaters in industry (+48\%) to produce industrial high-temperature heat.

\begin{figure}[t!]
\centering
\includegraphics[width=\textwidth]{Elec_Layer_REF.pdf}
\includegraphics[width=\textwidth]{Elec_Layer_SMR.pdf}
\caption{Balance between the production and the consumption of electricity in the REF (top) and SMR (down) scenarios. The production of electricity from \gls{SMR} substitutes more efficient technologies (\ie \gls{CCGT} and \gls{CHP}) and boosts the electrification of the rest of the system, mostly the industrial high-temperature heat sector.}
\label{fig:results_deter_layer_elec}
\end{figure}

@@

\newpage
Then, considering the primary energy mix shown in \Cref{fig:results_deter_energy_mix}, three phases in the transition can be identified. Before 2040, thanks to the perfect foresight approach, the model finds it more economical in 2025 to keep on using 33.2\,TWh of \gls{LFO} to produce \gls{HVC} through naphtha/LPG-cracking. In 2040, uranium-driven \gls{SMR} substitutes the electricity originally produced from industrial \gls{CHP} and \gls{CCGT} running respectively on fossil gas and renewable ammonia. Finally, from 2045 onward, the significant drop in the consumption of electrofuels comes from the same industrial \gls{CHP}. This is the illustration of the atom-molecules dilemma where the consumption of local renewables is, on its side, not much affected. In other words, \gls{SMR} competes with importing electrofuels while both support the integration of local solar and wind energies. Then, like the power sector, the SMR case ends up in a less efficient whole-energy system by 2050 at it consumes 47\,TWh (+12.7\%) primary energy more to supply an unchanged \gls{EUD}. Interestingly, in both cases, given the assumptions made on the \gls{GWP} of the resources (\ie $\mathit{gwp}_{\mathrm{op,electrofuels}}=0$\,kt$_{\ce{CO2},\text{eq}}$/GWh and $\mathit{gwp}_{\mathrm{op,uranium}}=0.004$\,kt$_{\ce{CO2},\text{eq}}$/GWh), the constraint on the \ce{CO2} budget leads to ``carbon neutrality'' by 2050\footnote{The model being constrained to keep on using all the waste that would keep on being locally produced, the system in 2050 reaches a $\sim 3.5\,Mt_{\ce{CO2},\text{eq}}$/year.}.

@@

@@

\newpage
As aforementioned, the industrial \gls{EUD} impacts the most the import of e-methane (Figure \ref{fig:results_uq_samples_gas}). This parameter directly dictates the demand of industrial high-temperature heat for which industrial gas \gls{CHP}, and industrial gas boilers to a lower extent, represent, on average over all the samples, respectively 25.6\% and 6.1\% of the total production. Then, considering the smaller-impact parameters, we notice that \gls{SMR} plays a non-negligible role. Indeed, if deployed, \gls{SMR} produces abundant low-emitting electricity for industrial electric heaters that substitute, even completely in some cases, gas alternatives. This confirms the observation made in Section \ref{subsec:atom_mol:results_deter_others}. Highly available local biomass also leads to smaller imports of e-methane to supply bio-hydrolysis and produce methane-equivalent gas. Finally, and surprisingly, the costs of purchasing electrofuels and fossil fuels have a positive and negative correlation with the amount of e-methane, respectively. In other words, by 2050, more expensive electrofuels result in more e-methane imports. On the contrary, cheaper electrofuels result in higher imports of fossil methane. Given the techno-economic optimum sought by EnergyScope, if electrofuels are more expensive, the system will, overall, import less of them, especially e-ammonia, mainly used by \gls{CCGT}. Subject to the \ce{CO2} budget for the transition, the system goes towards more efficient technologies, like industrial methane-\gls{CHP} to substitute e-ammonia-\gls{CCGT} in the production of electricity. First running on fossil gas, these \gls{CHP} consume more e-methane by 2050. On the contrary, if electrofuels are cheaper, there is more import of them, especially of e-ammonia. This leaves room for more emitting and cheaper resources to be used while respecting the \ce{CO2} budget, \ie coal in industrial boilers that produce, in these cases, on average 24\% of the high-temperature (HT) heat in 2050. In these cases, the use of coal in 2050 highlights that, with a sharper decrease of the emissions at early stages in the transition, the model finds a solution including highly-emitting resources (\eg coal) while respecting the \ce{CO2} budget. Consequently, there are smaller investments in methane \gls{CHP}, and consequently import of e-methane as more abundant renewable electricity is produced via e-ammonia-\gls{CCGT} and more HT-heat is supplied by industrial coal boilers. Even though we might expect that no more coal will be consumed in Belgium by 2050, the model still has the opportunity to use it if the \ce{CO2} budget allows it. Regarding the cost of purchasing fossil fuels, the parameter has mainly an impact on the import of fossil \gls{NG} as the most versatile energy carrier in the whole-energy system. If \gls{NG} is more expensive, the system will import less of it. Subsequently, the investments in methane-\gls{CHP} and boilers are more limited. This ends up in a smaller need for e-methane by 2050.

@@

\newpage
\section{Conclusions}
\label{sec:atom_mol:conclusions}
Given its lower \gls{LCOE} (40\,€/MWh$_{\text{elec}}$) and the important share of the investment recovered via its salvage value (between 79\% and 96\%), \gls{SMR} is installed as soon as available. It directly substitutes other flexible power generation units (\ie CCGT) and provides, by 2050, 44.6\,TWh, 25\% of the total electricity production. Consequently, this reduces the need to produce electricity via \gls{CHP} and allows increasing by 48\% the high-temperature heat produced by electric heaters. Besides these two sectors, the others (\ie low-temperature heat, mobility and non-energy demand) are marginally or not impacted by the integration of \gls{SMR}.

### 2024-09-02 chap_case_study/chap_case_study.tex
Ready to print

%\noindent
Works of \citet{rixhon2021comprehensive,rixhon2022integration} assessed the impact of the integration of the \gls{NED} in the case of Belgium in 2050, using EnergyScope TD (see Appendix \ref{app:ESTD}). To analyse the whole-energy system at different ``climate targets'', the model forced the total emissions to decrease by reducing their upper limit while optimising the total cost. In practice, 10\% steps of \gls{GWP} reduction were made from the ``reference scenario - 100\%''. This strategy gave the following points of analysis: 100\% (\ie cost-optimum with no limitation on the total \gls{GWP}), 90\%, 80\%, ..., down to 0\% (\ie  carbon neutrality).

@@

\citet{rixhon2021role} analysed the impact of these parameters on the total cost of the snapshot Belgian whole-energy system in 2050 subject to different \gls{GWP} limits. Based on this work, we have selected a subset of impacting uncertainties, added others due to the pathway formulation (\eg $\Delta_{\mathrm{change,pass}}$), and listed them in Table \ref{tab:UC_full}.

Table \ref{tab:UC_full} summarises the uncertainty ranges for the different groups of technologies and resources, for the year 2025.

>>>>

\newpage
Works of \citet{rixhon2021comprehensive,rixhon2022integration} assessed the impact of the integration of the \gls{NED} in the case of Belgium in 2050, using EnergyScope TD (see Appendix \ref{app:ESTD}). To analyse the whole-energy system at different ``climate targets'', the model forced the total emissions to decrease by reducing their upper limit while optimising the total cost. In practice, 10\% steps of \gls{GWP} reduction were made from the ``reference scenario - 100\%''. This strategy gave the following points of analysis: 100\% (\ie cost-optimum with no limitation on the total \gls{GWP}), 90\%, 80\%, ..., down to 0\% (\ie  carbon neutrality).

@@

\citet{rixhon2021role} analysed the impact of these parameters on the total cost of the snapshot Belgian whole-energy system in 2050 subject to different \gls{GWP} limits. Based on this work, we have selected a subset of impacting uncertainties, added others due to the pathway formulation (\eg $\Delta_{\mathrm{change,pass}}$), and listed them in Table \ref{tab:UC_full} that summarises the uncertainty ranges for the different groups of technologies and resources, for the year 2025.

### 2024-09-02 chap_methodo/chap_methodo.tex
Ready to print

\begin{figure}[htbp!]
\centering
\includegraphics[width=10cm]{MY_process_code.pdf}
\caption{Diagram of the iterative optimisation of the whole-energy system transition pathway. Myopic optimisation of the pathway consists of a sequence of smaller \gls{LP} problems limited to their respective 10-year time window. If the optimisation is not deterministic, step 3.3 is the one where the values of some parameters are changed due to uncertainties (see Section \ref{sec:meth:UQ}) or the actions taken by the Reinforcement Learning-agent (see Chapter \ref{chap:chap_RL})}
\label{fig:MY_process_code}
\end{figure}

@@

\newpage
Finally, the model accounts for thousands of parameters. The computational burden of considering all of them separately would be completely overwhelming ($\sim 10^7$ model runs). Similarly to other works \cite{Moret2017,limpens2020impact}, the model parameters that would follow the same uncertainty have been grouped into one single uncertain parameter. On top of mitigating the computational burden, this aims at grouping parameters that are closely linked with each other. For instance, the uncertainty on the cost of purchasing renewable electrofuels, $c_{\mathrm{op,electrofuels}}$, identically affects the cost of e-hydrogen, e-methane, e-ammonia and e-methanol. Indeed, besides their respective specificities, each of these fuels will be similarly affected by the variation in the cost of electricity or the electrolyser, which drive the majority of their cost of purchasing \cite{h2coalition}. Similarly, the uncertainties impacting the industrial demand, $industry\_\emph{EUD}$, alters equally the industrial high- and low-temperature and electricity demands as well as the non-energy demand. Section \ref{sec:cs:uncertainty} gathers the list of the uncertain parameters considered in this work.

@@

\newpage
Before jumping to the choice of the learning algorithm, it is worth noting that we opted for the combination of \gls{RL} with \gls{DNN}, called \gls{DRL}. Among others, one of the main drawbacks of traditional \gls{RL} algorithms, \ie without the use of \gls{NN},  is that it suffers from the ``curse of dimensionality'' when facing problems with continuous action and state spaces (see Chapter \ref{chap:chap_RL}). By approximating the state-action function with its parameters (\ie weights and biases), \gls{DNN} can address this difficulty.

@@

\gls{SAC} is a model-free and off-policy actor-critic deep \gls{RL} algorithm based on the entropy-augmented objective function (see Equation (\ref{eq:SAC_objective})). The word ``augmented'' here is in opposition to the conventional \gls{RL} objective function that is only based on the cumulative reward, \ie the first term of Equation (\ref{eq:SAC_objective}). In the \gls{RL} context, entropy, also called ``Shannon entropy'', stands for the randomness or stochasticity of the policy.

@@

From these observations, the main objective of \gls{PCA} is to find the linear combination, \ie projection, of the original variables that maximise their spread, \ie variance (see Figure \ref{fig:PCA_intro_proj}).  Another way to understand this is to look at the other side of the same coin. Indeed, \gls{PCA} also looks for properties that allow reconstructing the original features as accurately as possible. In practice, it aims at minimising the total reconstruction error which is the average squared distance between the original observations and their respective projection.

@@

\myparagraph{Outliers management}\\

@@

\newpage
\myparagraph{Robustness assessment of roadmaps}\\

>>>>

\begin{figure}[htbp!]
\centering
\includegraphics[width=\textwidth]{MY_process_code.pdf}
\caption{Diagram of the iterative optimisation of the whole-energy system transition pathway. Myopic optimisation of the pathway consists of a sequence of smaller \gls{LP} problems limited to their respective 10-year time window. If the optimisation is not deterministic, step 3.3 is the one where the values of some parameters are changed due to uncertainties (see Section \ref{sec:meth:UQ}) or the actions taken by the Reinforcement Learning-agent (see Chapter \ref{chap:chap_RL})}
\label{fig:MY_process_code}
\end{figure}

@@

Finally, the model accounts for thousands of parameters. The computational burden of considering all of them separately would be completely overwhelming ($\sim 10^7$ model runs). Similarly to other works \cite{Moret2017,limpens2020impact}, the model parameters that would follow the same uncertainty have been grouped into one single uncertain parameter. On top of mitigating the computational burden, this aims at grouping parameters that are closely linked with each other. For instance, the uncertainty on the cost of purchasing renewable electrofuels, $c_{\mathrm{op,electrofuels}}$, identically affects the cost of e-hydrogen, e-methane, e-ammonia and e-methanol. Indeed, besides their respective specificities, each of these fuels will be similarly affected by the variation in the cost of electricity or the electrolyser, which drive the majority of their cost of purchasing \cite{h2coalition}. Similarly, the uncertainties impacting the industrial demand, $industry\_\emph{EUD}$, alters equally the industrial high- and low-temperature and electricity demands as well as the non-energy demand. Section \ref{sec:cs:uncertainty} gathers the list of the uncertain parameters considered in this work.

@@

Before jumping to the choice of the learning algorithm, it is worth noting that we opted for the combination of \gls{RL} with \gls{DNN}, called \gls{DRL}. Among others, one of the main drawbacks of traditional \gls{RL} algorithms, \ie without the use of \gls{NN},  is that it suffers from the ``curse of dimensionality'' when facing problems with continuous action and state spaces (see Chapter \ref{chap:chap_RL}). By approximating the state-action function with its parameters (\ie weights and biases), \gls{DNN} can address this difficulty.

@@

\newpage
\gls{SAC} is a model-free and off-policy actor-critic deep \gls{RL} algorithm based on the entropy-augmented objective function (see Equation (\ref{eq:SAC_objective})). The word ``augmented'' here is in opposition to the conventional \gls{RL} objective function that is only based on the cumulative reward, \ie the first term of Equation (\ref{eq:SAC_objective}). In the \gls{RL} context, entropy, also called ``Shannon entropy'', stands for the randomness or stochasticity of the policy.

@@

\newpage
From these observations, the main objective of \gls{PCA} is to find the linear combination, \ie projection, of the original variables that maximise their spread, \ie variance (see Figure \ref{fig:PCA_intro_proj}).  Another way to understand this is to look at the other side of the same coin. Indeed, \gls{PCA} also looks for properties that allow reconstructing the original features as accurately as possible. In practice, it aims at minimising the total reconstruction error which is the average squared distance between the original observations and their respective projection.

@@

\newpage
\myparagraph{Outliers management}\\

@@

\myparagraph{Robustness assessment of roadmaps}\\

### 2024-09-02 frontend/merci.tex
Ready to print

% Promoteurs
Bien qu'un seul nom figure en couverture de cet ouvrage, les pages qui suivent seraient tristement restées vierges sans tout un environnement humain propice à la réflexion scientifique et à mon épanouissement personnel. Merci avant tout à mes deux promoteurs de thèse, mes deux ``pères académiques'', Hervé Jeanmart et Francesco Contino. Merci de m'avoir fait confiance dès le premier jour, donner la liberté de mener à bien mes recherches en toute autonomie tout en veillant à me prêter vos oreilles attentives et me prodiguer vos conseils avisés lors de nos échanges. Merci pour le temps que vous avez consacré lors de nos réunions pour résoudre les casse-têtes de ma thèse ainsi que ceux posés sur la table de Francesco. Merci à Hervé pour son pragmatisme et sa spontanéité et à Francesco pour son empathie et son goût du détail pour toujours chercher le dernier carat, à la limite parfois de la diptérophilie.

@@

% Collègues en commençant par Gauthier et Diederik
Merci ensuite à mes parrains, Gauthier et Diederik. Vous m'avez catapulté dans cette aventure et m'avez toujours accompagné pour explorer de nouveaux horizons. Merci à Marion et Ignace pour votre écolage dans l'univers du Reinforcement Learning. Enfin, outre les tontons (parfois) un peu (trop) grincheux, je ne saurais qu'assez remercier tous mes frères et soeurs/cousins et cousines de cette belle et grande famille du TFL. Dans le désordre, merci Alice pour ton humour et tes goûts musicaux ``discutables'' et avoir tenté tant bien que mal de m'initier à la science de la musique et à la culture flamande. Merci Arnaud pour ton amitié de longue date, ta disponibilité, tes relectures et tes opinions toujours magistralement construites. Merci coach PB de m'en avoir fait baver et de m'avoir poussé à dépasser mes limites lors de nos nombreuses courses. Merci Youyou d'y avoir supporté nos discussions de darons (et de me botter désormais les fesses). Merci Mamasita Aurel pour ta good vibe et ton eau pet', Martinos pour la co-création du label, Paolito pour ta force tranquille, Antoine aka Jabiru pour ton abonnement au MR magazine, p'tit Louis pour avoir toujours si bien veillé à ma roue arrière, Zakito pour ton grand coeur, Nicou pour ton flow et tes serrages. Merci Sara de ne pas me détester. Xie xie Yanan for the hot pot.

% Amis et famille
Je remercie aussi du fond du coeur ma famille, ma belle-famille et mes amis. Merci pour votre amour et votre indéfectible soutien qui m'ont permis d'avancer jusqu'ici et de devenir qui je suis.

% Chloé et les enfants
Last but not least, je remercie ma compagne, Chloé. Merci pour ton oreille attentive et bienveillante surtout dans mes moments d'errance quand j'envisageais une reconversion en tant que jardinier ou éleveur de chèvres en Suisse. Je te remercie aussi d'être celle qui m'accompagne au quotidien depuis toutes ces années, et, surtout, d'être la formidable maman de nos deux enfants, Victor et Oscar, et de me permettre ainsi d'exercer le plus beau métier du monde.

% Autres
Ci-dessus n'était que l'humble expression de ma profonde gratitude envers celleux qui m'ont accompagné tout au long de ce chemin tantôt long fleuve tranquille tantôt parcours semé d'embûches. J'ai certainement omis de mentionner nombreux d'entre vous. Sachez cependant que sans votre présence et votre soutien, ce beau projet professionnel mais, avant tout, cette fabuleuse aventure personnelle n'aurait pas été possible.

>>>>

% Promoteurs
Bien qu'un seul nom figure en couverture de cet ouvrage, les pages qui suivent seraient tristement restées vierges de lettres sans tout un environnement humain propice à la réflexion scientifique et à mon épanouissement personnel. Merci avant tout à mes deux promoteurs de thèse, mes deux ``pères académiques'', Hervé Jeanmart et Francesco Contino. Merci de m'avoir fait confiance dès le premier jour, donné la liberté de mener à bien mes recherches en toute autonomie tout en veillant à me prêter vos oreilles attentives et me prodiguer vos conseils avisés lors de nos échanges. Merci pour le temps que vous avez consacré lors de nos réunions afin de résoudre les casse-têtes de ma thèse ainsi que ceux posés sur la table de Francesco. Merci à Hervé pour ton pragmatisme et ta spontanéité et à Francesco pour ton empathie et ton goût du détail pour toujours chercher le dernier carat, à la limite parfois de la diptérophilie.

@@

% Collègues en commençant par Gauthier et Diederik
Merci ensuite à mes parrains, Gauthier et Diederik. Vous m'avez catapulté dans cette aventure et m'avez toujours accompagné pour explorer de nouveaux horizons. Merci à Marion et Ignace pour votre écolage dans l'univers du Reinforcement Learning. Enfin, outre les tontons (parfois) un peu (trop) grincheux, je ne saurais qu'assez remercier tous mes frères et soeurs/cousins et cousines de cette belle et grande famille du TFL. Dans le désordre, merci Alice pour ton humour et tes goûts musicaux ``discutables'' et avoir tenté tant bien que mal de m'initier à la science de la musique, à la culture flamande et au pastis. Merci Arnaud pour ton amitié de longue date, ta disponibilité, tes relectures et tes opinions toujours magistralement construites. Merci Coach PB de m'en avoir fait baver et de m'avoir poussé à dépasser mes limites lors de nos nombreuses courses. Merci Youyou d'y avoir supporté nos discussions de darons (et de me botter désormais les fesses). Merci Mamasita Aurel pour ta good vibe et ton eau pet', Martinos pour la co-création du label, Paolito pour ta force tranquille, Antoine aka Jabiru pour ton abonnement au MR magazine, p'tit Louis pour avoir toujours si bien veillé à ma roue arrière, Zakito pour ton grand coeur, Nicou pour ton flow et tes serrages. Merci Sara de ne pas me détester. Xie xie Yanan for the hot pot.

% Amis et famille
Je remercie aussi du fond du coeur ma famille, ma belle-famille et mes amis. Merci pour votre amour et votre indéfectible soutien qui m'ont permis d'avancer jusqu'ici et de devenir l'homme qui je suis aujourd'hui et que je deviendrai demain.

% Chloé et les enfants
Last but not least, je remercie ma compagne, ma Tchoup, Chloé. Merci pour ton oreille attentive et bienveillante surtout dans mes moments d'errance quand j'envisageais une reconversion en tant que jardinier ou éleveur de chèvres en Suisse. Je te remercie aussi d'être celle qui m'accompagne au quotidien depuis toutes ces années, et, surtout, d'être la formidable maman de nos deux enfants, Victor et Oscar, et de me permettre ainsi d'exercer le plus beau métier du monde.

% Autres
Ci-dessus n'était que l'humble expression de ma profonde gratitude envers celleux qui m'ont accompagné tout au long de ce chemin tantôt long fleuve tranquille tantôt parcours semé d'embûches. J'ai certainement omis de mentionner nombreux d'entre vous. Sachez cependant que sans votre présence et votre soutien, ce beau projet professionnel mais, avant tout, cette fabuleuse aventure personnelle n'auraient été possibles.
